A Spectral Analysis of the Domain Decomposed Adjoint Neumann-Ulam Method

### 2012-11-20 doc/mandc2013/notes/spectral_theory.tex
updated adjoint section and started on decomposition section in spectral note

%%---------------------------------------------------------------------------%%
\section{The Adjoint Neumann-Ulam Method}
We seek solutions of the general linear problem in the following form:
\begin{equation}
  \ve{A} \ve{x} = \ve{b}\:,
  \label{eq:linear_problem}
\end{equation}
where $\ve{A} \in \mathbb{R}^{N \times N}$ is a linear operator such
that $\ve{A} : \mathbb{R}^{N} \rightarrow \mathbb{R}^{N}$, $\ve{x} \in
\mathbb{R}^N$ is the solution vector, and $\ve{b} \in \mathbb{R}^N$ is
the forcing term. We choose the adjoint method Monte Carlo method to
invert the linear operator. We begin by defining the linear system
adjoint to Eq~(\ref{eq:linear_problem}):
\begin{equation}
  \ve{A}^T \ve{y} = \ve{d}\:,
  \label{eq:adjoint_linear_problem}
\end{equation}
where $\ve{y}$ and $\ve{d}$ are the adjoint solution and source
respectively and $\ve{A}^T$ is the adjoint operator. We can split this
equation by defining the following inner product equivalence
\cite{spanier_monte_1969}:
\begin{equation}
  \langle \ve{A}^T \ve{x}, \ve{y} \rangle = \langle \ve{x}, \ve{A}
  \ve{y} \rangle\:.
  \label{eq:adjoint_operator_product}
\end{equation}
With this statement we can then define the split equation:
\begin{equation}
  \ve{y} = \ve{H}^T \ve{y} + \ve{d}\:.
  \label{eq:adjoint_split_system}
\end{equation}
For convergence, we require that the spectral radius of $\ve{H}$ must
remain less than 1 as $\ve{H}^T$ contains the same eigenvalues and
therefore has the same spectral radius. From this definition it
follows that:
\begin{equation}
  \langle \ve{x}, \ve{d} \rangle = \langle \ve{y}, \ve{b} \rangle\:.
  \label{eq:adjoint_vector_relation}
\end{equation}
Using these definitions, we can derive an estimator from the adjoint
method that will also give the solution vector, $\ve{x}$. As with the
direct method, we can acquire the adjoint solution by forming the
Neumann series by writing Eq~(\ref{eq:adjoint_split_system}) as:
\begin{equation}
  \ve{y} = (\ve{I} - \ve{H}^T)^{-1} \ve{d}\:,
  \label{eq:adjoint_split_system_2}
\end{equation}
which in turn yields the Neumann series using the adjoint operator:
\begin{equation}
  \ve{y} = \sum_{k=0}^{\infty} (\ve{H}^T)^k\ve{d}\:.
  \label{eq:adjoint_neumann_series}
\end{equation}
We expand this summation to again yield a series of transitions that
can be approximated by a Monte Carlo random walk sequence, this time
forming the Neumann series in reverse order:
\begin{equation}
  y_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}h_{i_k,i_{k-1}}\ldots h_{i_2,i_1} h_{i_1,i} d_{i_k}\:.
  \label{eq:adjoint_neumann_solution}
\end{equation}
We can readily build an estimator for the adjoint solution from this
series expansion, but we instead desire the direct solution. We
achieve this by using Eq~(\ref{eq:adjoint_vector_relation}) as a
constraint. Here we have 2 unknowns, $\ve{y}$ and $\ve{d}$ and
therefore we require a second constraint to close the system. As a
second constraint we select:
\begin{equation}
  \ve{d} = \boldsymbol{\delta}_i\:,
  \label{eq:adjoint_second_constraint}
\end{equation}
where the $k^{th}$ component of the vector $\boldsymbol{\delta}_i$ is
the Dirac delta function $\delta_{i_k,i}$. If we apply
Eq~(\ref{eq:adjoint_second_constraint}) to our first constraint
Eq~(\ref{eq:adjoint_vector_relation}), we get the following convenient
outcome:
\begin{equation}
  \langle \ve{y}, \ve{b} \rangle = \langle \ve{x},
  \boldsymbol{\delta}_i \rangle = x_i \:,
  \label{eq:inner_product_constraint}
\end{equation}
meaning that if we compute the inner product of the direct source and
the adjoint solution using a delta function source, we recover the
direct solution we are looking for

@@

%%---------------------------------------------------------------------------%%
\section{Domain Decomposition}

>>>>

%%---------------------------------------------------------------------------%%
\section{The Adjoint Neumann-Ulam Method}
We seek solutions of the general linear problem in the following form:
\begin{equation}
  \ve{A} \ve{x} = \ve{b}\:,
  \label{eq:linear_problem}
\end{equation}
where $\ve{A} \in \mathbb{R}^{N \times N}$ is a linear operator such
that $\ve{A} : \mathbb{R}^{N} \rightarrow \mathbb{R}^{N}$, $\ve{x} \in
\mathbb{R}^N$ is the solution vector, and $\ve{b} \in \mathbb{R}^N$ is
the forcing term. Choosing the adjoint method Monte Carlo method to
invert the linear operator, we begin by defining the linear system
adjoint to Eq~(\ref{eq:linear_problem}):
\begin{equation}
  \ve{A}^T \ve{y} = \ve{d}\:,
  \label{eq:adjoint_linear_problem}
\end{equation}
where $\ve{y}$ and $\ve{d}$ are the adjoint solution and source
respectively and $\ve{A}^T$ is the adjoint operator.  With this
statement we can then define the split equation:
\begin{equation}
  \ve{y} = \ve{H}^T \ve{y} + \ve{d}\:,
  \label{eq:adjoint_split_system}
\end{equation}
where $\ve{H}$ is defined as the \textit{iteration matrix}.  For
convergence, we require that the spectral radius of $\ve{H}$ must
remain less than 1 as $\ve{H}^T$ contains the same eigenvalues and
therefore has the same spectral radius.  Using these definitions, we
can derive an estimator from the adjoint method that will also give
the solution vector, $\ve{x}$. As with the direct method, we can
acquire the adjoint solution by forming the Neumann series by writing
Eq~(\ref{eq:adjoint_split_system}) as:
\begin{equation}
  \ve{y} = (\ve{I} - \ve{H}^T)^{-1} \ve{d}\:,
  \label{eq:adjoint_split_system_2}
\end{equation}
which in turn yields the Neumann series using the adjoint operator:
\begin{equation}
  \ve{y} = \sum_{k=0}^{\infty} (\ve{H}^T)^k\ve{d}\:.
  \label{eq:adjoint_neumann_series}
\end{equation}
We expand this summation to again yield a series of transitions that
can be approximated by a Monte Carlo random walk sequence, this time
forming the Neumann series in reverse order:
\begin{equation}
  y_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}h_{i_k,i_{k-1}}\ldots h_{i_2,i_1} h_{i_1,i} d_{i_k}\:.
  \label{eq:adjoint_neumann_solution}
\end{equation}
We can readily build an estimator for the adjoint solution from this
series expansion, but we instead desire the direct solution. The
adjoint estimator can be related to the solution by defining the
following inner product equivalence \cite{spanier_monte_1969}:
\begin{equation}
  \langle \ve{A}^T \ve{x}, \ve{y} \rangle = \langle \ve{x}, \ve{A}
  \ve{y} \rangle\:.
  \label{eq:adjoint_operator_product}
\end{equation}
From this definition it follows that:
\begin{equation}
  \langle \ve{x}, \ve{d} \rangle = \langle \ve{y}, \ve{b} \rangle\:.
  \label{eq:adjoint_vector_relation}
\end{equation}
 Here we have 2 unknowns, $\ve{y}$ and $\ve{d}$ and therefore we
 require two constraints to close the system. We use
 Eq~(\ref{eq:adjoint_vector_relation}) as the first constraint and as
 a second constraint we select:
\begin{equation}
  \ve{d} = \boldsymbol{\delta}_i\:,
  \label{eq:adjoint_second_constraint}
\end{equation}
where the $k^{th}$ component of the vector $\boldsymbol{\delta}_i$ is
the Dirac delta function $\delta_{i_k,i}$. If we apply
Eq~(\ref{eq:adjoint_second_constraint}) to our first constraint
Eq~(\ref{eq:adjoint_vector_relation}), we get the following convenient
outcome:
\begin{equation}
  \langle \ve{y}, \ve{b} \rangle = \langle \ve{x},
  \boldsymbol{\delta}_i \rangle = x_i \:,
  \label{eq:inner_product_constraint}
\end{equation}
meaning that if we compute the inner product of the direct source and
the adjoint solution using a delta function source, we recover the
direct solution we are looking for

@@

%%---------------------------------------------------------------------------%%
\section{Domain Decomposition}
For very large linear systems, two things are likely to occur: a
single compute node does not contain enough memory to hold the entire
problem or a single compute node takes a prohibitively long time to
solve the system. Domain decomposition methods permit the global
computational domain to be subdivided into smaller regions. These
smaller regions require less memory to store and fewer computational
resources to solve as compared to the global problem. Typically, these
smaller regions are then represented separately in indiviudal compute
nodes. Therefore, in order to solve large scale problems with the
Neumann-Ulam method, we must explore its behavior in a domain
decomposed evironment.

For reactor physics Monte Carlo simulations, domain decomposition has
been identified as a key principle in moving forward in high
performance computing
\cite{brunner_comparison_2006,siegel_analysis_2012}. To accomplish
this, we recognize from the literature that stochastic histories must
be transported from domain to domain as the simulation progress and
they transition to states that are not in the local domain. Because we
have chosen a domain decomposition strategy in a parallel environment,
this means that communication of these histories must occur between
compute nodes owning neighboring pieces of the global domain.
